---
title: "Identifying genes related to response using drug screen metrics"
output: html_document
date: "2026-01-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	fig.width = 5,
	fig.height=5
)

```

```{r}
library(pheatmap)
library(ggplot2)
library(edgeR)
library(limma)
library(DT)
library(readxl)
library(RColorBrewer)
```

```{r}
### Functions 

### buttons for datatables 
create_dt <- function(x){
  DT::datatable(x,
                extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                               buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                               lengthMenu = list(c(10,25,50,-1),
                                                 c(10,25,50,"All"))))
}
```

## Notes

The aim of this analysis is to use relative AUC values (generated from dose response curves of Olivia's organoid drug screens), as a metric of sensitivity/response, to identify genes related to response for diffferent chemotherapeutic drugs. 

relAUC was calculated from dose response curves using the MESS package in R (cell viability vs log10(conc)), and the following cut offs were defined to assess response: 

• 0 < relAUC < 0.5 -> SENSITIVE 

• 0.5 < relAUC < 0.75 -> INTERMEDIATE

• relAUC > 0.75 -> RESISTANT 

We require at least n=3 for each response group within a drug to compare differences in gene expression between response. 

DATA: 

• RNAseq data :'/home/ndalvi/RNAseq_STAR/Results/HTseq/collated_counts.tsv'

• AUC data:'/data/AUC_metadata_JAN2026.xlsx'


## Data {.tabset}

Taking a quick look at both the RNA-seq data and AUC data 

RNA-seq data:
```{r}
rawdata <- read.delim2('/Volumes/bioinf/home/ndalvi/RNAseq_STAR/Results/HTseq/collated_counts.tsv', header = TRUE, check.names = FALSE, stringsAsFactors = FALSE)
subset <- c('ENSEMBL_GeneID','ORG-38-ORG','ORG41','ORG-46-ORG','ORG49','ORG55','ORG60','ORG61','ORG64_new','ORG65','ORG66','ORG70','ORG73')
head(rawdata[,subset] )
```

Relevant metadata:

### relAUC values
```{r}
metadata_continuous <- read_excel("/Users/dalvinikita/Documents/7_GitHub/drug-AUC-analysis/data/AUC_metadata_JAN2026.xlsx", sheet = 'continuous')
metadata_continuous <- metadata_continuous[-c(5,14,15),]
create_dt(metadata_continuous)
```

### Sensitivity Class

```{r}
metadata_discrete <- read_excel("/Users/dalvinikita/Documents/7_GitHub/drug-AUC-analysis/data/AUC_metadata_JAN2026.xlsx", sheet = 'discrete')
metadata_discrete <- metadata_discrete[-c(5,14,15),]
create_dt(metadata_discrete)
```


## Method 1: Correlative Analysis 

For the correlative analysis, we use the relAUC values for each drug to create a design matrix and fit a GLM to our gene expression data, to identify genes related to differences in relAUC for a particular drug. 

### Design 
Taking a look at the design matrix which includes growth pattern, batch of sequencing, and dose:

```{r}

### creating a DGElist object to include organoids with AUC data
organoids <- c('ORG-38-ORG','ORG41','ORG-46-ORG','ORG49','ORG55','ORG60','ORG61','ORG64_new','ORG65','ORG66','ORG70','ORG73')
y<- DGEList(counts=rawdata[,organoids], genes=rawdata[,1])

### tidying up sample names
colnames(y) <- c('ORG38-O','ORG41','ORG46-O','ORG49','ORG55','ORG60','ORG61','ORG64','ORG65','ORG66','ORG70','ORG73')

### tidying up metadata sample names to match counts matrix]
metadata_continuous$PATIENT <- c('ORG38-O','ORG41','ORG46-O','ORG49','ORG55','ORG60','ORG61','ORG64','ORG65','ORG66','ORG70','ORG73')
rownames(metadata_continuous) <- metadata_continuous$PATIENT

### creating factors to add into design matrix
### including pattern, batch, and dose
pattern <- as.factor(metadata_continuous$PATTERN)
batch <- as.factor(metadata_continuous$BATCH)
dose <- as.factor(metadata_continuous$DOSE)

#create design matrix to include batch, growth pattern, and dose
design <- model.matrix(~0+pattern+batch+dose)
rownames(design) <- metadata_continuous$PATIENT
create_dt(design)
```

### Normalisation

Filtering to remove genes with low counts:

```{r, fig.width=7}
#filtering to remove low counts
keep <- filterByExpr(y, design)
print(table(keep))
y <- y[keep, , keep.lib.sizes=TRUE]

#Visualiszing library sizes pre and post normalisation
col <- brewer.pal(32, 'Paired')

par(mfrow=c(1,2))
lcpm <- cpm(y, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")

y <- calcNormFactors(y) 
lcpm <- cpm(y, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```

### MDS plots {.tabset}

```{r, echo=FALSE}
col.pattern <- c('darkorchid3', 'deeppink')[as.factor(metadata_continuous$PATTERN)]
col.batch <- c('darkorchid3', 'deeppink', 'orange')[as.factor(metadata_continuous$BATCH)]
col.dose <- c('darkorchid3', 'deeppink')[as.factor(metadata_continuous$DOSE)]
```

#### Label
```{r, fig.height=10, fig.width=10}
par(mfrow=c(2,2))
plotMDS(y, col=col.pattern)
title(main="A. Growth Pattern")
legend("topleft",
       fill=c('darkorchid3', 'deeppink'), 
       legend=levels(as.factor(metadata_continuous$PATTERN)))

plotMDS(y, col=col.batch)
title(main="B. Batch")
legend("topleft",
       fill=c('darkorchid3', 'deeppink','orange'), 
       legend=levels(as.factor(metadata_continuous$BATCH)))

plotMDS(y, col=col.dose)
title(main="C. DOSE")
legend("topleft",
       fill=c('darkorchid3', 'deeppink'), 
       legend=levels(as.factor(metadata_continuous$DOSE)))

```

#### Point
```{r, fig.height=10, fig.width=10}
par(mfrow=c(2,2))
plotMDS(y, col=col.pattern, pch=19)
title(main="A. Growth Pattern")
legend("topleft",
       fill=c('darkorchid3', 'deeppink'),
       legend=levels(as.factor(metadata_continuous$PATTERN)))

plotMDS(y, col=col.batch, pch=19)
title(main="B. Batch")
legend("topleft",
       fill=c('darkorchid3', 'deeppink','orange'),
       legend=levels(as.factor(metadata_continuous$BATCH)))

plotMDS(y, col=col.dose, pch = 19)
title(main="C. DOSE")
legend("topleft",
       fill=c('darkorchid3', 'deeppink'), 
       legend=levels(as.factor(metadata_continuous$DOSE)))
```

### Analysis {.tabset}

In this approach, we run 7 separate analyses (for  drugs except Staurosporine, 5FU, Carboplatin, Cisplatin, and Oxaliplatin - which do not show sufficient variation in response) to identify genes associated with changes in AUC values for the respective drug, across our 12 organoid lines. Design matrices for all analysis will include: 

• Growth pattern (INF/EXP)

• Batch of sequencing (1,2,3)

• AUC values for respective drug as a numeric/ continuous variable. 

• Doses of drug received (1 or 2)

#### Docetaxel

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata_continuous$Docetaxel)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

Top genes related to Docetaxel AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf)
create_dt(as.data.frame(genes))
```
*Notes: 0 genes found to have FDR < 0.05*

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]
auc_values <- auc

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

#### Doxrubicin

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata_continuous$Doxorubicin)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

Top genes related to Doxrubicin AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf)
create_dt(as.data.frame(genes))
```
*Notes: 0 genes found to have FDR < 0.05*

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]
auc_values <- auc

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

#### Gemcitabine

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata_continuous$Gemcitabine)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

Top genes related to Gemcitabine AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf)
create_dt(as.data.frame(genes))
```
*Notes: 0 genes found to have FDR < 0.05*

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]
auc_values <- auc

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

#### Irinotecan

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata_continuous$Irinotecan)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

Top genes related to Irinotecan AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf)
create_dt(as.data.frame(genes))
```
*Notes: 0 genes found to have FDR < 0.05*

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]
auc_values <- auc

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

#### Mitomycin

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata_continuous$Mitomycin)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

Top genes related to Mitomycin AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf)
create_dt(as.data.frame(genes))
```
*Notes: 5 genes found to have FDR < 0.05*

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]
auc_values <- auc

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

#### Paclitaxel

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata_continuous$Paclitaxel)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

Top genes related to Paclitaxel AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf)
create_dt(as.data.frame(genes))
```
*Notes: 0 genes found to have FDR < 0.05*

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]
auc_values <- auc

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

#### Topotecan

Design matrix used for analysis:

```{r}
counts <- y
auc <- as.numeric(metadata_continuous$Topotecan)
auc_design <- cbind(design,auc)
create_dt(auc_design)
```

Top genes related to Topotecan AUC: 

```{r}
counts <- counts %>%
  estimateDisp() %>%
  estimateCommonDisp() %>%
  estimateTagwiseDisp()

fit <- glmQLFit(counts, design = auc_design, robust = TRUE)
auc_fit <- glmQLFTest(fit, coef=c('auc'))
genes <- topTags(auc_fit, n=Inf)
create_dt(as.data.frame(genes))
```
*Notes: 0 genes found to have FDR < 0.05*

Plotting top genes: 

```{r}
top_genes <- c(genes$table$genes)
top_genes <- top_genes[1:5]
auc_values <- auc

for (gene in top_genes) {
  gene <- as.character(gene)
  y$cpm <- cpm(y, log = T) 
  expression_values <- y$cpm[y$genes$genes==gene,]
  plot<- ggplot(data = NULL, mapping = aes(x=auc, y=expression_values, colour = batch)) + 
    geom_point() +
    theme_linedraw() +
    xlab('AUC values') + 
    ylab('log(cpm)') +
    ggtitle(label=gene)
  print(plot)
}
```

## Method 2A: Differential Expression by Drug 

```{r}
rawdata <- read.delim2('/Volumes/bioinf/home/ndalvi/RNAseq_STAR/Results/HTseq/collated_counts.tsv', header = TRUE, check.names = FALSE, stringsAsFactors = FALSE)
subset <- c('ENSEMBL_GeneID','ORG-38-ORG','ORG41','ORG-46-ORG','ORG49','ORG55','ORG60','ORG61','ORG64_new','ORG65','ORG66','ORG70','ORG73')
```

*Note: We have removed ORG76 from the AUC metadata as there is no corresponding RNAseq, and for organoids with AUC data for 1 dose AND 2 dose experimental methods, we keep data only from the 1 dose regimen for a more balanced dataset.*

### QC and Normalisation

#### Design 
Taking a look at the design matrix which includes growth pattern, batch of sequencing, drug sensitivity classification, and dose:

```{r}
### creating a DGElist object to include organoids with AUC data
organoids <- c('ORG-38-ORG','ORG41','ORG-46-ORG','ORG49','ORG55','ORG60','ORG61','ORG64_new','ORG65','ORG66','ORG70','ORG73')
y<- DGEList(counts=rawdata[,organoids], genes=rawdata[,1])

### tidying up sample names
colnames(y) <- c('ORG38-O','ORG41','ORG46-O','ORG49','ORG55','ORG60','ORG61','ORG64','ORG65','ORG66','ORG70','ORG73')

### tidying up metadata sample names to match counts matrix]
metadata_discrete$PATIENT <- c('ORG38-O','ORG41','ORG46-O','ORG49','ORG55','ORG60','ORG61','ORG64','ORG65','ORG66','ORG70','ORG73')
rownames(metadata_discrete) <- metadata_discrete$PATIENT

### creating factors to add into design matrix
### including pattern and batch
pattern <- as.factor(metadata_discrete$PATTERN)
batch <- as.factor(metadata_discrete$BATCH)
dose <- as.factor(metadata_discrete$DOSE)
response <- as.factor(metadata_discrete$Mitomycin)

#create design matrix to include batch, growth pattern
design <- model.matrix(~0+response+pattern+batch+dose)
rownames(design) <- metadata_discrete$PATIENT
create_dt(design)
```

#### Normalisation

Filtering to remove genes with low counts:

```{r, fig.width=7}
#filtering to remove low counts
keep <- filterByExpr(y, design)
print(table(keep))
y <- y[keep, , keep.lib.sizes=TRUE]

#Visualiszing library sizes pre and post normalisation
col <- brewer.pal(32, 'Paired')

par(mfrow=c(1,2))
lcpm <- cpm(y, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")

y <- calcNormFactors(y) 
lcpm <- cpm(y, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```

### Differential Expression 

To ensure adequate power, we require at least 3 samples in each group for comparison (i.e n=3 resistant, and n=3 sensitive). This requirement is only satisfied for Mitomycin: 

```{r}
table(metadata_discrete$Mitomycin)
```

Hence, we can look at the difference in expression between organoids resistant to mitomycin and those sensitive to mitomycin, while accounting for dose, sequencing batch, and growth pattern: 

```{r}
create_dt(design)

### fit data to GLM
fit <- glmQLFit(y, design, robust=TRUE)

### create contrast
resVSsens <- makeContrasts(responseResistant - responseSensitive, levels = colnames(design))
res <- glmQLFTest(fit, contrast=resVSsens)
DEgenes <- topTags(res, n=Inf)
create_dt(DEgenes$table)

```


## Method 2B: Differential Expression by Drug Class

```{r}
rawdata <- read.delim2('/Volumes/bioinf/home/ndalvi/RNAseq_STAR/Results/HTseq/collated_counts.tsv', header = TRUE, check.names = FALSE, stringsAsFactors = FALSE)
subset <- c('ENSEMBL_GeneID','ORG-38-ORG','ORG41','ORG-46-ORG','ORG49','ORG55','ORG60','ORG61','ORG64_new','ORG65','ORG66','ORG70','ORG73')
metadata_discrete <- read_excel("/Users/dalvinikita/Documents/7_GitHub/drug-AUC-analysis/data/AUC_metadata_JAN2026.xlsx", sheet = 'discrete')
metadata_discrete <- metadata_discrete[-c(5,14,15),]
```

*Note: We have removed ORG76 from the AUC metadata as there is no corresponding RNAseq, and for organoids with AUC data for 1 dose AND 2 dose experimental methods, we keep data only from the 1 dose regimen for a more balanced dataset.*

### QC and Normalisation

#### Design 

Taking a look at the design matrix which includes growth pattern, dose, drug class response, and batch of sequencing:

```{r}
### creating a DGElist object to include organoids with AUC data
organoids <- c('ORG-38-ORG','ORG41','ORG-46-ORG','ORG49','ORG55','ORG60','ORG61','ORG64_new','ORG65','ORG66','ORG70','ORG73')
y<- DGEList(counts=rawdata[,organoids], genes=rawdata[,1])

### tidying up sample names
colnames(y) <- c('ORG38-O','ORG41','ORG46-O','ORG49','ORG55','ORG60','ORG61','ORG64','ORG65','ORG66','ORG70','ORG73')

### tidying up metadata sample names to match counts matrix]
metadata_discrete$PATIENT <- c('ORG38-O','ORG41','ORG46-O','ORG49','ORG55','ORG60','ORG61','ORG64','ORG65','ORG66','ORG70','ORG73')
rownames(metadata_discrete) <- metadata_discrete$PATIENT

### creating factors to add into design matrix
### including pattern and batch
pattern <- as.factor(metadata_discrete$PATTERN)
batch <- as.factor(metadata_discrete$BATCH)
dose <- as.factor(metadata_discrete$DOSE)
response <- as.factor(metadata_discrete$Cytotoxic_Antibiotics)

#create design matrix to include batch, growth pattern
design <- model.matrix(~0+response+pattern+batch+dose)
rownames(design) <- metadata_discrete$PATIENT
create_dt(design)
```

#### Normalisation
Filtering to remove genes with low counts:
```{r, fig.width=7}
#filtering to remove low counts
keep <- filterByExpr(y, design)
print(table(keep))
y <- y[keep, , keep.lib.sizes=TRUE]

#Visualiszing library sizes pre and post normalisation
col <- brewer.pal(32, 'Paired')

par(mfrow=c(1,2))
lcpm <- cpm(y, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")

y <- calcNormFactors(y) 
lcpm <- cpm(y, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```

### Differential Expression 

To ensure adequate power, we require at least 3 samples in each group for comparison (i.e n=3 resistant, and n=3 sensitive). This requirement is only satisfied for the Cytotixic Antibiotics drug class: 

```{r}
table(metadata_discrete$Cytotoxic_Antibiotics)
```

Hence, we can look at the difference in expression between organoids resistant to mitomycin and those sensitive to mitomycin, while accounting for dose, sequencing batch, and growth pattern: 

```{r}
create_dt(design)

### fit data to GLM
fit <- glmQLFit(y, design, robust=TRUE)

### create contrast
resVSsens <- makeContrasts(responseResistant - responseSensitive, levels = colnames(design))
res <- glmQLFTest(fit, contrast=resVSsens)
DEgenes <- topTags(res, n=Inf)
create_dt(DEgenes$table)

```

Three DE genes with FDR < 0.05: 

• SAA1

• TFF1

• TSHZ2
